Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-18 src/tex/ms.tex
[showyourwork] overleaf sync

Signatures of habitability, from ocean glint to optical effects from condensates --- like the rainbow---should be some of the most distinct signals in an exoplanet's phase curve, particularly in polarized light.
%
However, depending on system architectures, some of these features can be obscured by the inner working angle of a coronagraph.

Surface liquid water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet therefore provides important context for interpreting possible biosignatures.
%
Providing additional context to the habitability of an exoplanet is the presence of a water-cycle as evidenced by the characterization of a rainbow and other cloud optical effects.

@@

An exoplanet observed in an edge-on orbit will move through all phases and thus all scattering angles, while those viewed at higher inclinations will pass through a subset of these scattering angles.
%
If the planet is rotating, this supplies additional regular variations that can be used to place constraints on the heterogeneity of the planet and potentially map it.

The strengths of detailed phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scatter, rainbows, and glories.
%
In polarized light an added potential benefit is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is from the specular (Fresnel) reflection of light off of smooth liquid surfaces at crescent phases.
%
In the total flux the signal of the glint will be most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at 53 degrees  \citep[a scattering angle of 127 degrees; see e.g. ][]{2008Icar..195..927W}.
%
It therefore varies slightly by wavelength and by sea and atmosphere composition.
%
In general on a planet the glint will be strongest in red wavelengths where the Rayleigh contribution to both the phase curve and attenuation are minimized \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also minimized \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of surface liquid bodies, which taken in the context of the temperature and condensates may be interpreted as the presence of liquid water.
%
The forward scattering from clouds can be confused with glint \citep{Robinson_2010} further motivating also probing the rainbow angle for context.
%
The cloud coverage fraction on ocean worlds can also be constrained with observations spanning the ultraviolet to near-infrared by the point phase curves at all wavelengths intersect in polarized light \citep{treesandstam2019}.
%
Additionally, \citet{trees2022} showed that glint can also be detected in spectropolarimetry while the planet is at phase angles placing it at the wide separations more easily probed with coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, glories and other optical effects from condensates provide a wealth of information about an exoplanet \citep{karalidi2012rainbow, stam2008}.
%
In our own Solar System the polarized back scatter from Titan provided evidence that the world's reflectance was due to haze rather than a surface \citep{zellner1973polarization}, on Venus the rainbow and phase curve behavior with wavelength showed the clouds were $\sim$75\% H$_2$SO$_4$ -- water mixture also constraining the droplet size distribution \citep{hansenhovenier1974}, and the glory was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

>>>>

Signatures of habitability, from ocean glint to optical effects from condensates --- like the rainbow--- should be some of the most distinct signals in an exoplanet's reflected light phase curve, particularly in polarization.
%
However, depending on system architectures, some of these features can be obscured by the inner working angle of a coronagraph.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and e.g.\ geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.
%
Providing additional context to the habitability of an exoplanet is the presence of a water-cycle as evidenced by the identification of a rainbow and other optical effects that can be attributed to water clouds.

@@

An exoplanet observed in an edge-on orbit will move through all phases and thus all scattering angles, while those viewed at higher inclinations will pass through a subset of these scattering angles.
%
If the planet is rotating, its reflected light signal is expected to show regular variations that can place constraints on the horizontal heterogeneity of the planet.

The strengths of detailed phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface, 2022A&A...664A..59M}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scatter, rainbows, and glories.
%
An added potential benefit of observing polarized light is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade as these stars themselves can be considered to be unpolarized \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces.
%
In the total flux the glint signal will be most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at 53$^\circ$ \citep[a scattering angle of 127 degrees; see e.g. ][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with wavelength and with sea and atmosphere composition.
%
In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
{\color{red}{check the next sentence:}}
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of liquid surface bodies, which taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
The forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent phases, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry when the planet is at phase angles where it is at the wide separations more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, glories and other optical effects from condensates provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,2014A&A...566L...1G}.
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to haze rather than a surface \citep{zellner1973polarization}, on Venus the rainbow and phase curve behavior with wavelength showed the clouds were $\sim$75\% H$_2$SO$_4$ -- water mixture also constraining the droplet size distribution \citep{hansenhovenier1974}, and the glory was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.
